Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-12-06 draft_lm1.tex
fixing stuff; beginning methods

% Our defs 
\def \rr {$R_{t}\ $}

@@

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease.

@@

The attack ratio is also influenced by the reproductive number of the disease 
\cite{bacaer_final_2009, katriel_attack_2012}. With Dengue, since transmission 
is affected by environmental conditions as discussed above, the reproductive 
number is also expected to vary with time.

Various challenges exist to assess the attack rate of a dengue epidemic. One of 
them is that there are a large number of asymptomatic cases in any epidemic, 
which nevertheless acquire immunity. It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years. Due to the alternation of 
ciculating serotypes, we detected fluctuating levels of susceptibility the 
circulating virus(es). To Estimate the susceptibles we have we have fixed the 
tramissibility based on the $R_t$ series derived from data\cite{nishiura}. Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

To accomplish the estimation of the $S_0$, we use a simplified model of 
dengue transmission based on a single-strain Susceptible-infecctious-Removed 
(SIR) model. The reason for the choice of a single strain model, is the lack of 
accurate information regarding the actual proportions of each virus in 
circulation. However, some information is available about the predominant 
serotypes for some epidemics in the period 
of study\cite{macedo_virological_2013}.

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}. These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?} In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

The authors have applied similar methods as used here, to estimate the number 
of susceptibles for the predominant strain of influenza at the beginning of  
influenza seasons in europe\cite{pone2011}.

@@

In this section we will start by describing how we estimated the effective 
reproductive number, $R_t$, relative to the global dengue cases, and its 
posterior probability distribution. 
We then proceed to describe the SIR model 
used to represent the global incidence and how the $R_t$ is integrated into the 
model to allow for time varying transmission. 
Then we describe the Bayesian 
estimation methodology applied to find the posterior probability distributions 
of the susceptibles to the main circulating Dengue viruses for each epidemic, 
using Dengue incidence time series for Rio de Janeiro from 1996 to 2014.
In the last subsection we describe how we estimated the Attack ratio of each 
epidemic from the estimated susceptibility and observed incidence.

\subsection*{Estimating the effective reproductive number, $R_t$}

In monitoring of infectious diseases, it is important to assess whether the 
incidence of a  particular disease is increasing significantly, in order to 
decide to take preventive measures.
The effective reproductive number at time $t$, \rr, can be understood as a 
real-time estimate of the basic reproductive number ($R_{0}$) and is defined as 
the average number of secondary cases per primary case at time $t$.
Here we derive confidence/credibility bounds for the uncertainty about the 
estimate of the effective reproductive number $R_t$ under the assumption that 
reported disease counts are Poisson distributed.
It is shown that one can construct a conservative confidence interval for \rr 
using the induced probability distribution that results from the approach of 
\cite{mantel}.

Let $Y_t$ be the number of reported disease cases for a particular time $t \in 
(0, T)$.
\cite{nishiura} extend the theory developed by \cite{stallybrass} and propose 
to estimate \rr as
\begin{equation}
\label{eq:Rtestimate}
R_t = \left( \frac{Y_{t+1}}{Y_t}\right)^{1/n}
\end{equation}
where $n$ is taken to be the ratio between the length of reporting interval and 
the mean generation time of the disease.
Here we are interested in the simpler case $n=1$.
If \rr is to be used as a decision tool, one needs to be able to quantify the 
uncertainty about estimate in equation~\ref{eq:Rtestimate}. 
Here we detail how to obtain confidence intervals for \rr, exploring two 
approaches under the assumption that the counts $Y_t$ are Poisson distributed 
for all $t$.

First we explore the approach of \cite{mantel}, whose objective is to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors then proceed to note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
\cite{mantel} show that one can construct a $100\alpha \%$ confidence interval 
for \rr by noting that
\begin{align}
 R_t^{(L)} &= \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})}\\
 R_t^{(U)} &= \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}
\end{align}

>>>>

% Our defs 
\def \rr {$R_{t}\:$}

@@

Dengue transmission is modulated by environmental conditions, of which 
temperature, due to its effects on the 
vector reproduction, stands out as a strong predictor of 
incidence~\cite{honorio_temporal_2009,wu_higher_2009}.
In places with sufficient temperature variation dengue is predominantly a 
summer disease.

@@

The attack ratio is also influenced by the reproductive number of the disease 
~\cite{bacaer_final_2009, katriel_attack_2012}.
With Dengue, since transmission 
is affected by environmental conditions as discussed above, the reproductive 
number is also expected to vary with time.

Various challenges exist to assess the attack rate of a dengue epidemic. 
One of 
them is that there are a large number of asymptomatic cases in any epidemic, 
which nevertheless acquire immunity.
It is estimated that for 
every case reported, 10-20 are not seen by health authorities (personnal 
communication).
Thirdly, under-reporting is a serious issue in Brazil. 
Duarte and Fran\c{c}a (2006)~\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting for hospitalized patients in 
Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the suspected dengue cases go unreported.  
Lastly, demography and migrations affect the number of susceptible in ways which 
are not easy to determine.

In this paper, we estimate the number of susceptibles before every major 
outbreak ($S_{0, year}$) in the last 18 years.
Due to the alternation of 
ciculating serotypes, we detected fluctuating levels of susceptibility the 
circulating virus(es).
To estimate the susceptibles we have we have fixed the 
tramissibility based on the $R_t$ series derived from data~\cite{nishiura}. 
Then, 
from the incidence series and the population at risk, we could then calculate 
the attack ratio for each epidemic.

To accomplish the estimation of the $S_0$, we use a simplified model of 
dengue transmission based on a single-strain Susceptible-Infecctious-Removed 
(SIR) model.
The reason for the choice of a single strain model, is the lack of 
accurate information regarding the actual proportions of each virus in 
circulation.
However, some information is available about the predominant 
serotypes for some epidemics in the period 
of study~\cite{macedo_virological_2013}.

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?}
In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

The authors have applied similar methods as used here, to estimate the number 
of susceptibles for the predominant strain of influenza at the beginning of  
influenza seasons in europe~\cite{pone2011}.

@@

In this section we will start by describing a method to estimate the 
effective 
reproductive number, $R_t$, and obtain its 
posterior distribution. 
We then proceed to describe Susceptible-infectious-recovered (SIR) model 
used to represent the global disease incidence and how $R_t$ can be integrated 
into the model to allow for time varying transmission. 
Next, an approach to approximate the posterior distributions 
of the numbers of susceptible to the main circulating Dengue viruses for each 
epidemic is detailed.
Finally, we discuss how to estimate the attack ratio of each 
epidemic using the estimated susceptible fraction and observed incidence.
We employ these methods to analyze Dengue incidence time series for the city of 
Rio de Janeiro from 1996 to 2014.

\subsection*{Estimating the effective reproductive number ($R_t$)}

 In monitoring of infectious diseases, it is important to assess whether the 
incidence of a  particular disease is increasing significantly, in order to 
decide to take preventive measures.
The effective reproductive number at time $t$, \rr, can be understood as a 
real-time estimate of the basic reproductive number ($R_{0}$) and is defined as 
the average number of secondary cases per primary case at time $t$.

Let $Y_t$ be the number of reported disease cases for a particular time $t \in 
(0, T)$.
Nishiura el al. (2010)~\cite{nishiura} extend the theory developed by 
Stallybrass et al. (1931)~\cite{stallybrass} and propose to estimate \rr as
\begin{equation}
\label{eq:Rtestimate}
R_t = \left( \frac{Y_{t+1}}{Y_t}\right)^{1/n}
\end{equation}
where $n$ is taken to be the ratio between the length of reporting interval and 
the mean generation time of the disease.
Here we are interested in the simpler case $n=1$.
If \rr is to be used as a decision tool, howeve, one needs to be able to 
quantify the 
uncertainty about estimate in equation~\ref{eq:Rtestimate}. 
Here we detail how to obtain credibility intervals for \rr under the assumption 
that the counts $Y_t$ are Poisson distributed for all $t$.

We explore the approach of Ederer and Mantel~\cite{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors then proceed to note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
\cite{mantel} show that one can construct a $100\alpha \%$ confidence interval 
for \rr by noting that
\begin{align}
 R_t^{(L)} &= \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})}\\
 R_t^{(U)} &= \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}
\end{align}
